Emittance preservation of an electron beam in a loaded quasi-linear plasma wakefield

### 2017-09-12 Tex/Paper.tex
Cleanup

% Math
\newcommand{\unit}[1]{\,\mathrm{#1}}
\newcommand{\funit}[2]{\,\mathrm{#1}/\mathrm{#2}}
\newcommand{\mexp}[1]{\mathrm{e}^{#1}}
\newcommand{\nexp}[1]{\times 10^{#1}}

@@

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativist beam travelling through a plasma
excites a strong longitudinal e-field that can be loaded by a trailing witness beam. With optimal beam loading, increase
in energy spread can be kept to a minimum. Acceleration of an electron withness beam by an electron drive beam has been
demonstrated experimentally \cite{rosenzweig:1988, blumenfeld:2007, kallos:2008} in the past. AWAKE at CERN is a proof
of concept proton driven plasma wakefield accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, the non-linear transverse wakefields cause emittance growth of the accelerated
beam. A finite length beam will also see a transversely varying accelerating field causing increasing energy spread
\cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by the transverse
oscillations of the plasma electrons forming a sheet around an evacuated area leaving only ions. The ions, assumed
stationary, form an ion channel creating a focusing force that varies linearly with radius. This produces an axially
symmetric focusing force \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
%~ \begin{align}
    %~ \lambda_{p} = 2\pi\frac{c}{\omega_{pe}} \label{EQ:LambdaP}
%~ \end{align}
%~ and
%~ \begin{align}
    %~ \omega_{pe} = \sqrt{\frac{n_{0} e^{2}}{m_{e} \epsilon_{0}}}, \label{EQ:OmegaP}
%~ \end{align}
$\lambda_{p} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$,
will produce a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness beam loading the
peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches. Acceleration of
an electron witness beam driven by two electron drive beams was demonstrated at Brookhaven National Laboratory
\cite{muggli:2011}.

The energy carried by electron drive bunches used in previous experiments were typically small, on the order of
$100\unit{J}$, and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. The energy of a
high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam in the international linear collider with
$1\mexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers as drivers a large number of plasma stages
is required. How to efficiently stage plasma accelerators is challenging without reducing the effective gradient and
spoiling the beam quality \cite{steinke:2016,lindstrom:2016}. Proton drive beams available at CERN carry
significantly more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for longer accelerator
stages. However, the SPS beam is orders of magnitude longer than the plasma wavelengths needed for such applications.
This issue is resolved by letting the proton beam undergo self-modulation before injecting the witness beam into one of
the buckets in the modulated structure. This self-modulation instability is caused by the transverse fields generated by
the beam acting upon the beam itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation
frequency is close to that of the plasma, and leaves a train of proton bunches along the beam axis with a surrounding
halo.
%~ De-phasing of the
%~ witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
%~ beam, limits the effective length of a plasma accelerator stage.

@@

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver, and a nominal plasma density $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam
such that $k_{pe}\sigma_{x} = 1$ where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first
phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the
wakefield with a long electron beam. The study presented here is for Run 2 \cite{adli:2016}, which aims to demonstrate
acceleration of a short electron beam to high energy with a minimal increase in emittance and energy spread.

@@

The preliminary design of the second run proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without
the electron beam present. The electron witness beam will be injected into the modulated proton beam before stage two,
where acceleration will occur.
%~ As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
%~ keep this as short as possible.

With this experimental design the self-modulated proton beam does not produce a non-linear wakefield, and not all plasma
electrons are therefore evacuated. The result is that the focusing force will not increase linearly with radius. The
small size of the accelerating structure produced by the self-modulated beam also put constraints on the size of the
electron witness beam, and thus its charge and current, if we wish to prevent large energy spread. By matching the
transverse size to a practical emittance and the plasma density, we prevent large amplitude oscillations which may cause
additional growth in energy spread as well as in emittance.

@@

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors
that may affect this, we have tried several approaches to create a stable drive beam structure based on previous self-
modulation studies.

@@

In order to study the witness beam emittance evolution we turned to the recently released open source version of
QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D quasi-static PIC code \cite{huang:2006, an:2013}.
Since QuickPIC is a quasi-static code, it does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, it is well suited to study emittance preservation.

@@

In these simulations we use a single proton drive beam that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in the second plasma cell of AWAKE Run 2 (Figure \ref{Fig:AWAKER2}). When the
initial SPS beam containing $3\nexp{11}$ protons \cite{gschwendtner:2016} enters the second plasma cell the peak $E_{z}$
field is expected to be $500-600\funit{MV}{m}$. As the baseline AWAKE drive beam current is insufficient to reach the
non-linear regime and produce a plasma bubble, the plasma electrons are only depleted to around $65\%$ of nominal plasma
density at the point where we inject the electron beam \cite{awake_collaboration:2016}. These conditions are replicated
in our simulations using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and $7\unit{kA}$.

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. In order to create a stable environment to study the
evolution of the witness beam we prevented the proton beam from evolving radially by increasing the particle mass by a
factor of $1\mexp{6}$. We are not considering the evolution of the proton beam itself in this study.

@@

The witness beam in our simulation differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. The relative drift causing the de-phasing $\Delta\xi \propto \gamma^{-2}$. By simply matching the
initial $\gamma$ values we eliminate this effect entirely, but a lower initial value is likely to be sufficient
\cite{berglyd_olsen:2015}.

In these 3D simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{x}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_{pe}}{2\pi}\sqrt{2\gamma_{rel}}, \label{EQ:Matched}
\end{align}
where $\lambda_{p}$ is the plasma wavelength. This is a very narrow beam compared to the drive beam
$\sigma_{x} = 200\unit{\mu m}$. The charge density of this compact witness beam reaches that of the plasma density at
only a few $\unit{pC}$, but reaches optimal beam loading at around $1-200\unit{pC}$. The implication here being that the
witness beam produces its own non-linear wakefield already at the head of the beam. The majority of the electrons
within it will therefore see a linear focusing force preventing the bulk of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam. We used a transverse grid cell size of $1.17\unit{\mu m}$, and $2.34\unit{\mu m}$ for
the longitudinal grid cells. The witness beam was simulated with $16.8\unit{M}$ non-weighted particles.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} The top plot compares the unloaded longitudinal e-field (no witness beam, blue.
        dashed  line) and the loaded e-field (blue line) along the beam axis. The magnitude of the beam density along
        the axis is shown (red line) for reference. The bottom plot compares the plasma density along the beam axis for
        a drive beam with no witness beam (dashed green line), witness beam with no drive beam (dash-dotted green line),
        and both witness and drive beam present (continuous green line). $\xi = z - tc$ is the position in the
        simulation box and both beams travel towards the left. The beam and plasma densities arein units of initial
        plasma density $n_{0} = 7\nexp{-14}\unit{cm}^{-3}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} The plasma density in grey with the proton beam (blue) and the electron beam
        (red) superimposed. The line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis.}
\end{figure}

@@

The witness beam generates its own wakefield which partially cancels out the $E_{z}$ field generated by the drive beam.
With an ideally shaped electron beam charge profile it is possible to optimally load the field, such that the
accelerating $E_{z}$ is constant along the bunch \cite{katsouleas:1987, tzoufras:2009}.

@@

\begin{figure*}[hbt]
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittance}
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittanceOffset}
    \caption{\label{Fig:BeamEmitt} The red lines show a moving window calculation of transverse normalised emittance for
        an on-axis beam with respect to the drive beam axis (left) and an offset beam (right) with an offset of one
        $\sigma_{x} = 5.24\unit{\mu m}$ in the x plane. The moving window for emittance calculation is longitudinal
        slices of $l = 4\times\Delta\xi = 18.75\unit{\mu m}$ with a $\Delta\xi$ resolution. Only slices with more than
        $100$ macro particles have been included, which accounts for the abrupt edge as well as the noise at either end
        of the beam due to low statistics. The blue lines show the peak electron beam density profile. For reference,
        the plasma density profile is included in green, but scaled up by a factor of $100$ to be visible. The solid
        lines are at the plasma entry point, the dashed lines after $4\unit{m}$ of plasma, the dash-dotted after
        $40\unit{m}$, and the dotted after $100\unit{m}$. In order to sustain a stable accelerating field for the
        witness beam, these simulations were run with an LHC energy drive beam of $7\unit{TeV}$ to avoid de-phasing
        which causes the structure to break down after about $50\unit{m}$.}
\end{figure*}

@@

While the ideal shape is trapezodial, for Gaussians beams a reasonable flat field, and consequently low energy spread,
can be achieved by controlling the charge density to achieve a similar effect. This, however, puts the front of the
witness beam outside of the ideal region while the middle bulk of the beam has a trapezoidal-like shape. In our case,
with a matched beam to the plasma density defined by equation \ref{EQ:Matched}, we are limited on how much total charge
we can accelerate without reaching charge densities that will overload the field.

As illustrated in figure \ref{Fig:BeamPS} our base case shows a difference in energy transfer to the tail of the beam
compared to the centre where the $E_{z}$-field is nearly flat. Also the front of the beam has a large tail in momentum
space as this regions sees a smaller accelerating field. The positioning of the witness beam is chosen to put the bulk
of the charge as close to the peak accelerating field as possible.

The charge density required to optimally load the $E_{z}$ field is $n_{eb} \approx 35\times n_{0}$. This means that the
witness beam's own wakefield is in the fully non-linear regime, where the space charge force is sufficient to blow-out
all plasma electrons resulting in an ion column. This ion column, as is well known \cite{rosenzweig:1991}, provides a
linear focusing force on the part the electron beam within the ion column, and therefore prevents emittance growth for
this region. This effect is shown for our base case in figure \ref{Fig:PlasmaDenTWake}. The focusing field has a
gradient of $20\unit{kT/m}$ near the beam axis.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied that maintained a stable accelerating structure, about $70-80\%$ of the beam
retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows emittance growth for the base case as a function of
propagation length. No significant emittance growth was observed for propagation lengths up to $100\unit{m}$. The drive
beam energy increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing. De-phasing for the SPS beam case starts to
become significant after about $50\unit{m}$.

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius follows the evolution given by equation \ref{EQ:Matched} with $\lambda_{pe} = \lambda_{0}$ for the section
of the beam where normalised emittance is preserved as shown in figure \ref{Fig:BeamFilament}. This has the potential to
cause overloading of the field. However, for our base case this effect is not significant.

@@

Since the accelerated electron beam creates its own focusing bubble, the emittance of the part of the beam inside the
bubble is not affected by small beam offsets with respect to the proton beam axis. This is an added benefit of this new
accelerating regime, and may ease the transverse injection tolerances. The head of the beam does not benefit from this
effect, but since the proton beam creates a quasi linear wake, the head of the beam still stabilises after some time.
This is illustrated on figure \ref{Fig:BeamEmitt} for an electron beam offset of $1\sigma_{x}$. In this case there is a
larger initial emittance growth (see left hand plots of figure \ref{Fig:BeamFilament}), but the emittance growth stop
after the first few metres (figure \ref{Fig:BeamEmitt}). This effect is likely to be greater for larger offsets as the
beam oscillates around the axis of the drive beam wakefield (see right hand plots of figure \ref{Fig:BeamFilament}).

@@

Figures \ref{Fig:BeamQ} and \ref{Fig:BeamQAbs} show the dependence on bunch length and bunch charge for an initial beam
emittance of $2\unit{\mu m}$ after propagating through $4\unit{m}$ of plasma. We ran the scan with beam charges from
$10\unit{pC}$ to $300\unit{pC}$ and with length $\sigma_{z}$ from $40\unit{\mu m}$ to $100\unit{\mu m}$. We see from
figure \ref{Fig:BeamQ} that both the $40\unit{\mu m}$ and the $60\unit{\mu m}$ beam has a well defined minimum energy
spread with a beam charge $\geq 50\unit{pC}$ and $\approx 100\unit{pC}$ respectively. Lower beam charges tend to
underload the $E_{z}$-field, while higher beam charges tend to overload it. It is also clear that long beams with
respect to the accelerating flank of the $E_{z}$-field, $\approx\lambda_{pe}/4$, will not optimally load the field along
its length thus experiencing a larger spread in energy.

@@

Figure \ref{Fig:BeamQEmit} shows how the growth in emittance and energy spread varies with initial electron beam
emittance. The smaller the initial emittance is, the better the emittance is preserved. There are two effects that lead
to growth for high emittance beams; the transverse beam size may increase beyond the size of the bubble, or the beam
density may be reduced so much that the plasma electrons are no longer fully evacuated from the ion column. Emittance
values higher than a few micrometers lead to a significant increase in both emittance and energy spread. In these
simulations the focusing of the beam to the plasma remains matched for each emittance value.

>>>>

% Math
\newcommand{\unit}[1]{\,\mathrm{#1}}
\newcommand{\funit}[2]{\,\mathrm{#1}/\mathrm{#2}}
\newcommand{\mexp}[1]{\mathrm{e}{#1}}
\newcommand{\nexp}[1]{\times 10^{#1}}

@@

Beam driven plasma wakefield accelerators have the potential to offer compact linear accelerators with high energy
gradients, and have been of interest for several decades \cite{chen:1985}. A relativistic beam travelling through a
plasma excites a strong longitudinal e-field that can be loaded by a trailing witness beam. With optimal beam loading,
an increase in energy spread can be kept to a minimum. Acceleration of an electron withness beam by an electron drive
beam has been demonstrated experimentally in the past \cite{rosenzweig:1988, blumenfeld:2007, kallos:2008}. AWAKE at
CERN is a proof of concept proton driven plasma wakefield accelerator experiment \cite{awake_collaboration:2014}.

A major challenge with plasma wakefield accelerators is, however, to produce an accelerated beam while keeping growth in
energy spread and emittance as low as possible. In the well described linear case where the beam density $n_{b}$ is much
smaller than the plasma density $n_{0}$, the non-linear transverse wakefields cause emittance growth of the accelerated
beam. The beam will also see a transversely varying accelerating field causing increasing energy spread
\cite{katsouleas:1987}. In the non-linear regime, where $n_{b} > n_{0}$, a bubble is formed by the transverse
oscillations of the plasma electrons forming a sheet around an evacuated area leaving only ions. The ions, assumed
stationary, form an ion channel creating a focusing force that varies linearly with radius. This produces an axially
symmetric focusing force \cite{lu:2006-1, lu:2006}.
%~ This effect also avoids the issue of increasing energy spread due to transverse beam loading. Increase in energy spread
%~ is still dependent on beam loading $E_{z}$.
%~ Ideal beam loading can be achieved by specially shaped beams that optimally flattens the field
%~ \cite{katsouleas:1987, tzoufras:2009}, but this is difficult experimentally and is not a topic explored in this paper.

@@

A train of electron drive bunches with a separation $\lambda_{pe}$ and a length $l_{b} \ll \lambda_{pe}$, where
%~ \begin{align}
    %~ \lambda_{p} = 2\pi\frac{c}{\omega_{pe}} \label{EQ:LambdaP}
%~ \end{align}
%~ and
%~ \begin{align}
    %~ \omega_{pe} = \sqrt{\frac{n_{0} e^{2}}{m_{e} \epsilon_{0}}}, \label{EQ:OmegaP}
%~ \end{align}
$\lambda_{p} = 2\pi c/\omega_{pe}$ and $\omega_{pe} = (n_{0} e^{2} / m_{e} \epsilon_{0})^{1/2}$,
will produce a field $E_{z}$ that increases for each drive bunch \cite{chen:1985}. A tailing witness beam loading the
peak accelerating phase of this field will gain energy from the wakefield from all of the drive bunches. Acceleration of
an electron witness beam driven by two electron drive beams has been demonstrated at Brookhaven National Laboratory
\cite{muggli:2011}.

The energy carried by electron drive bunches used in previous experiments were typically small -- on the order of
$100\unit{J}$ -- and the propagation length typically $<1\unit{m}$ \cite{blumenfeld:2007,caldwell:2009}. The energy of a
high-charge electron beam accelerated to $1\unit{TeV}$, similar to the beam in the international linear collider with
$1\nexp{10}$ electrons, is $1.6\unit{kJ}$. By using electron beams or lasers as drivers, a large number of plasma stages
is required. How to efficiently stage plasma accelerators is challenging without reducing the effective gradient and
spoiling the beam quality \cite{steinke:2016,lindstrom:2016}. Proton drive beams available at CERN carry significantly
more energy, $19\unit{kJ}$ for the SPS beam \cite{gschwendtner:2016}, allowing for longer accelerator stages. However,
the SPS beam is orders of magnitude longer than the plasma wavelengths needed for such applications. This issue is
resolved by letting the proton beam undergo self-modulation before injecting the witness beam into one of the buckets in
the modulated structure. This self-modulation instability is caused by the transverse fields generated by the beam
acting upon the beam itself, causing regions of the beam to rapidly defocus \cite{kumar:2010}. The modulation frequency
is close to that of the plasma, and leaves a train of proton bunches along the beam axis with a surrounding halo.
%~ De-phasing of the
%~ witness beam relative to the drive beam, caused by loss of energy in the drive beam and gain of energy in the witness
%~ beam, limits the effective length of a plasma accelerator stage.

@@

%~ The first run of the experiment is set up with a $10\unit{m}$ rubidium vapour cell.
%~ The rubidium vapour is
%~ ionised by a $780\unit{nm}$, $4.5\unit{TW}$ peak power laser with a pulse length of $100−120\unit{fs}$.
The AWAKE experiment, currently in its first stages of operation at CERN, uses the SPS beam at $400\unit{GeV}$ as its
driver, and a nominal plasma density $7\nexp{14}\unit{cm}^{-3}$ \cite{gschwendtner:2016}. This plasma density
corresponds to a plasma wavelength $\lambda_{pe} = 1.26\unit{mm}$. The plasma density is matched to the SPS proton beam
such that $k_{pe}\sigma_{x} = 1$, where $k_{pe} = 2\pi/\lambda_{pe}$ is the plasma wave number. The aim of the first
phase of Run 1 of the experiment is to demonstrate self-modulation of the proton beam, and in 2018 to sample the
wakefield with a long electron beam. The study presented here is for Run 2 \cite{adli:2016}, which aims to demonstrate
acceleration of a short electron beam to high energy with a minimal increase in emittance and energy spread.

@@

The preliminary design of Run 2 proposes to use two plasma sections, illustrated in figure \ref{Fig:AWAKER2}.
The first section of $4\unit{m}$ is the self-modulation stage where the proton beam undergoes self-modulation without
the electron beam present. The electron witness beam will then be injected into the modulated proton beam before stage
two, where it undergoes acceleration.
%~ As the $E_z$-field will decrease due to the gap between the two cells, it is desirable to
%~ keep this as short as possible.

With this experimental design the self-modulated proton beam does not produce a full non-linear wakefield, and not all
plasma electrons are therefore evacuated. The result is that the focusing force will not increase linearly with radius.
The small size of the accelerating structure produced by the self-modulated beam also puts constraints on the size of
the electron witness beam, and thus its charge and current -- if we at the same time wish to prevent large energy
spread. By matching the transverse size of the beam at a given emittance to the plasma density, we prevent large
amplitude oscillations which may cause additional growth in energy spread as well as in emittance.

@@

The main focus of this study has been on the beam loading of the electron beam. In order to eliminate other factors
that may affect this (add examples?), we have tried several approaches to create a stable drive beam structure based on
previous self-modulation studies.

@@

In order to study the witness beam emittance evolution we turned to the recently released open source version of
QuickPIC developed by UCLA. QuickPIC is a fully relativistic 3D quasi-static PIC code \cite{huang:2006, an:2013}.
Since QuickPIC is a quasi-static code, it does not suffer from the numerical Cherenkov effect that full PIC codes do
\cite{godfrey:1974,lehe:2013}, making it a well suited code to study emittance preservation.

@@

In these simulations we use a single proton drive beam that sets up a wakefield comparable to that which we expect to
see from the self-modulated SPS beam in the second plasma cell of AWAKE Run 2 (Figure \ref{Fig:AWAKER2}). When the
initial SPS beam containing $3\nexp{11}$ protons \cite{gschwendtner:2016} enters the second plasma cell, the peak
$E_{z}$ field is expected to be $500-600\funit{MV}{m}$. As the baseline AWAKE drive beam current is insufficient to
reach the non-linear regime and produce a plasma bubble, the plasma electrons are only depleted to around $65\%$ of
nominal plasma density at the point where we inject the electron beam \cite{awake_collaboration:2016}. These conditions
are replicated in our simulations using a single proton beam of $1.46\nexp{10}$ protons, or $2.34\unit{nC}$ and
$7\unit{kA}$.

The peak density of the simulation drive beam is $0.83\cdot n_{0}$, producing a quasi-linear wakefield. The single bunch
setup uses the baseline proton energy $W_{pb} = 400\unit{GeV}$ and transverse size $\sigma_{x} = 200\unit{\mu m}$. We
also set the length of the drive beam to $\sigma_{z} = 40\unit{\mu m}$. In order to create a stable environment to study
the evolution of the witness beam we in addition prevented the proton beam from evolving radially by increasing the
particle mass by a factor of $1\mexp{6}$. We are not considering the evolution of the proton beam itself in this study.

@@

The witness beam in our simulations differs from AWAKE baseline parameters on several key points. Initial beam energy is
set such that $\gamma_{eb} = \gamma_{pb} = 426.3$. This was done to eliminate the problem of initial de-phasing of the
witness beam. Beam loading of a short witness beam is sensitive to its position relative to the field
\cite{tzoufras:2009}. The relative drift causing the de-phasing is proportional to $\gamma^{-2}$. By simply matching the
initial $\gamma$ values we eliminate this effect entirely, but a lower initial value is likely to be sufficient
\cite{berglyd_olsen:2015}.

In these 3D simulations we have used a matched beam for an initial emittance of $2\unit{\mu m}$, corresponding to a
$\sigma_{x}$ of $5.25\unit{\mu m}$. The beam matching relation is given by
\begin{align}
    \beta = \frac{\sigma_r^2}{\epsilon_g} = \frac{\lambda_{pe}}{2\pi}\sqrt{2\gamma}, \label{EQ:Matched}
\end{align}
where $\lambda_{pe}$ is the plasma wavelength and $\gamma$ is the beam relativistic factor. This relation requires a
very narrow beam compared to the drive beam $\sigma_{x} = 200\unit{\mu m}$. The charge density of this compact witness
beam reaches that of the plasma density at only a few $\unit{pC}$, but reaches optimal beam loading at around
$1-200\unit{pC}$. The implication here being that the witness beam produces its own non-linear wakefield already at the
head of the beam. The majority of the electrons within it will therefore see a linear focusing force preventing the bulk
of the beam from undergoing emittance growth.

The relatively small size of the witness beam compared to the proton beam put some restrictions on the transverse
resolution of the witness beam. We used a transverse grid cell size of $1.17\unit{\mu m}$, and $2.34\unit{\mu m}$ for
the longitudinal grid cells. The witness beam was simulated with $16.8\unit{M}$ non-weighted particles. For the largest
parameter scans, shown in figure \ref{Fig:BeamQ}, we used a transverse grid cell size of $2.34\unit{\mu m}$.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamLoading}
    \caption{\label{Fig:BeamLoading} The top plot compares the unloaded longitudinal e-field (no witness beam, blue.
        dashed line) and the loaded e-field (blue line) along the beam axis. The magnitude of the beam density along
        the axis is shown (red line) for reference. The bottom plot compares the plasma density along the beam axis for
        a drive beam with no witness beam (dashed green line), witness beam with no drive beam (dash-dotted green line),
        and both witness and drive beam present (continuous green line). $\xi = z - tc$ is the position in the
        simulation box, and both beams travel towards the left. The beam and plasma densities are in units of initial
        plasma density $n_{0} = 7\nexp{-14}\unit{cm}^{-3}$.}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/plasmaDenTWake}
    \caption{\label{Fig:PlasmaDenTWake} The plasma density in grey with the proton beam (blue) and the electron beam
        (red) superimposed. The super-imposed line plot indicates the transverse wakefield gradient $dW_{x}/dx$ where
        $W_{x} = E_{x} - v_{b} B_{y}$, evaluated along the beam axis.}
\end{figure}

@@

The witness beam generates its own wakefield which partially cancels out the $E_{z}$ field generated by the drive beam.
With an ideally shaped electron beam charge profile it is possible to optimally load the field in such a way that the
accelerating $E_{z}$ is constant along the bunch \cite{katsouleas:1987, tzoufras:2009}.

@@

\begin{figure*}[hbt]
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittance}
    \includegraphics[width=0.495\linewidth,trim={2mm 0mm 2mm 0mm},clip]{figures/beamEmittanceOffset}
    \caption{\label{Fig:BeamEmitt} The red lines show a moving window calculation of transverse normalised emittance for
        an on-axis beam with respect to the drive beam axis (left), and an offset beam (right) with an offset of one
        $\sigma_{x} = 5.24\unit{\mu m}$ in the x-plane. The moving window for emittance calculation is longitudinal
        slices of $l = 4\times\Delta\xi = 9.38\unit{\mu m}$ with a $\Delta\xi$ resolution. Only slices with more than
        $100$ macro particles have been included. The blue lines show the on-axis electron beam density profile. For
        reference, the plasma density profile is included in green, but scaled up by a factor of $100$ to be visible.
        The solid lines are taken at the plasma entry point, the dashed lines after $4\unit{m}$ of plasma, the
        dash-dotted after $40\unit{m}$, and the dotted after $100\unit{m}$. In order to sustain a stable accelerating
        field for the witness beam, these simulations were run with an LHC energy drive beam of $7\unit{TeV}$ to avoid
        de-phasing which otherwise causes the structure to break down after about $50\unit{m}$ of plasma.}
\end{figure*}

@@

While the ideal shape is trapezoidal, for Gaussian beams a reasonable flat field, and consequently low energy spread,
can be achieved by controlling the charge density to achieve a similar effect. This, however, puts the front of the
witness beam outside of the ideal region while the middle bulk of the beam has a trapezoidal-like shape. In our case,
with a matched beam to the plasma density defined by equation \ref{EQ:Matched}, we are limited on how much total charge
we can accelerate without reaching charge densities that will overload the field.

As illustrated in figure \ref{Fig:BeamPS} our base case shows a difference in energy transfer to the tail of the beam
compared to the centre where the $E_{z}$-field is nearly flat. In addition, the front of the beam has a large tail in
momentum space as this regions sees a smaller accelerating field. The positioning of the witness beam is chosen to put
the bulk of the charge as close to the peak accelerating field as possible.

The charge density required to optimally load the $E_{z}$ field is $n_{eb} \approx 35\times n_{0}$. This means that the
witness beam's own wakefield is in the fully non-linear regime, where the space charge force is sufficient to blow-out
all plasma electrons resulting in an ion column. This ion column, as is well known \cite{rosenzweig:1991}, provides a
linear focusing force on the part the electron beam within the column, and therefore prevents emittance growth for this
region of the beam. This effect is shown for our base case in figure \ref{Fig:PlasmaDenTWake}. The focusing field has a
gradient of $20\unit{kT/m}$ near the beam axis.

@@

%~ The witness beam's self-generated bubble prevents the tailing end of the beam to see any significant emittance growth.
For the majority of the cases we studied that maintained a stable accelerating structure, about $70-80\%$ of the beam
retained its initial emittance. Figure \ref{Fig:BeamEmitt} shows emittance growth for the base case as a function of
propagation length. No significant emittance growth was observed for propagation lengths up to $100\unit{m}$. The drive
beam energy was here increased to $7\unit{TeV}$ (LHC energy) to prevent de-phasing, as de-phasing starts to become a
significant effect for the SPS beam after about $50\unit{m}$.

The on-axis density of the electron beam increases as its gamma factor increases and its transverse size decreases. The
beam radius follows the evolution given by equation \ref{EQ:Matched} with $\lambda_{pe} = \lambda_{0}$ for the section
of the beam where normalised emittance is preserved -- as shown in figure \ref{Fig:BeamFilament}. This has the potential
to cause overloading of the field. However, for our base case this effect was not significant.

@@

Since the accelerated electron beam creates its own focusing bubble, the emittance of the part of the beam inside the
bubble is not affected by small beam offsets with respect to the proton beam axis. This is an added benefit of this new
accelerating regime, and may ease the transverse injection tolerances. The head of the beam does not benefit from this
effect, but since the proton beam creates a quasi linear wake, the head of the beam still stabilises after some time.
This is illustrated on figure \ref{Fig:BeamEmitt} for an electron beam offset of $1\sigma_{x}$. In this case there is a
larger initial emittance growth (see left hand plots of figure \ref{Fig:BeamFilament}), but the emittance growth stop
after the first few metres (figure \ref{Fig:BeamEmitt}). This effect is likely to be greater for larger offsets as the
beam oscillates around the axis of the drive beam wakefield, as can bee seen from the right hand plots of figure
\ref{Fig:BeamFilament}.

@@

Figures \ref{Fig:BeamQ} and \ref{Fig:BeamQAbs} show the dependence on bunch length and bunch charge for an initial beam
emittance of $2\unit{\mu m}$ after propagating through $4\unit{m}$ of plasma. We ran the scan with beam charges from
$10\unit{pC}$ to $300\unit{pC}$ and with length $\sigma_{z}$ from $40\unit{\mu m}$ to $100\unit{\mu m}$. We see from
figure \ref{Fig:BeamQ} that both the $40\unit{\mu m}$ and the $60\unit{\mu m}$ beam has a well defined minimum energy
spread with a beam charge $\geq 50\unit{pC}$ and $\approx 100\unit{pC}$ respectively. Lower beam charges tend to
underload the $E_{z}$-field, while higher beam charges tend to overload it. It is also clear that long beams with
respect to the accelerating flank of the $E_{z}$-field, $\approx\lambda_{pe}/4$, will not optimally load the field along
its length, thus producing a larger spread in energy.

@@

Figure \ref{Fig:BeamQEmit} shows how the growth in emittance and energy spread varies with initial electron beam
emittance. The smaller the initial emittance is, the better the emittance is preserved. There are two effects that lead
to growth for high emittance beams; the transverse beam size may increase beyond the size of the bubble, or the beam
density may be reduced so much that the plasma electrons are no longer fully evacuated from the ion column. Emittance
values higher than a few micrometres lead to a significant increase in both emittance and energy spread. In these
simulations the focusing of the beam to the plasma remains matched for each emittance value.
